---
title: "1. Introduction: Offline, Causal, and Online Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Introduction: Offline, Causal, and Online Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `rLifting` package provides a unified high-performance framework for Wavelet Lifting Transforms in R. Uniquely, it bridges the gap between traditional **offline statistical analysis** and **real-time signal processing**.

This vignette demonstrates the three core modes of operation using a Doppler signal:

1.  **Offline (Batch):** global denoising using future information.
2.  **Causal (Batch):** batched processing that respects time causality (no lookahead).
3.  **Online (Stream):** point-by-point processing for real-time applications.

## Setup and Data

We use the `doppler_example` dataset included in the package.

```{r setup}
library(rLifting)
library(ggplot2)

# Load example data (Doppler signal with Gaussian noise)
data("doppler_example", package = "rLifting")

head(doppler_example)
```

## 1. Offline Denoising (Global)

In **Offline** mode, the algorithm has access to the entire signal (past and future). This allows it to compute global thresholding statistics and perform non-causal lifting steps, resulting in the smoothest possible reconstruction.

**Use case:** Historical data analysis, post-processing.

```{r offline}
# Define the wavelet scheme (Haar in this example)
scheme <- lifting_scheme("haar")

# Perform Global Denoising
# Extension="symmetric" is standard to reduce boundary artifacts.
offline_clean <- denoise_signal_offline(
  doppler_example$noisy,
  scheme,
  levels = 3, # You can choose the analysis resolution level
  method = "semisoft",
  extension = "symmetric"
)

# Plot
df_plot <- doppler_example
df_plot$Offline <- offline_clean

ggplot(df_plot, aes(x = index)) +
  geom_line(aes(y = noisy), color = "grey80") +
  geom_line(aes(y = original, color = "Truth"), linewidth = 0.8) +
  geom_line(aes(y = Offline, color = "Offline"), linewidth = 0.8) +
  scale_color_manual(values = c("Truth" = "black", "Offline" = "blue")) +
  theme_minimal() +
  labs(title = "Offline Denoising", y = "Amplitude")
```

## 2. Causal Denoising (Batch)

In **Causal** mode, we process the history as if we were receiving it sequentially. The estimate at time $t$ depends *only* on $x_1, \dots, x_t$. We use a sliding window of size $W$ to calculate local statistics (see Liu et al 2014).

**Use case:** Backtesting trading strategies, simulating real-time systems on historical data.

```{r causal}
# Causal Denoising
# window_size = 256 means the transform considers only the last 256 points at each step.
causal_clean <- denoise_signal_causal(
  doppler_example$noisy,
  scheme,
  window_size = 256,
  levels = 3,
  method = "semisoft",
  extension = "symmetric"
)

df_plot$Causal <- causal_clean

# Notice the "Warm-up" phase at the beginning where the signal is 0 or raw
ggplot(df_plot, aes(x = index)) +
  geom_line(aes(y = original, color = "Truth"), alpha = 0.5) +
  geom_line(aes(y = Causal, color = "Causal"), linewidth = 0.8) +
  scale_color_manual(values = c("Truth" = "black", "Causal" = "red")) +
  theme_minimal() +
  labs(title = "Causal Denoising (Simulated Real-Time)", y = "Amplitude")
```

## 3. Online Denoising (Stream)

**Online** mode allows you to process data point-by-point as it arrives from a sensor or API. It uses a `C++` Ring Buffer to maintain the sliding window state efficiently ($O(1)$ per update).

**Use case:** IoT sensors, live financial feeds, monitoring systems.

```{r stream}
# 1. Initialize the stream processor
processor <- new_wavelet_stream(
  scheme,
  window_size = 256,
  levels = 3,
  method = "semisoft",
  extension = "symmetric"
)

# 2. Simulate a stream loop
stream_output <- numeric(nrow(doppler_example))

# In a real app, this loop would be receiving data from a socket/API
for (i in 1:nrow(doppler_example)) {
  new_point <- doppler_example$noisy[i] 
  stream_output[i] <- processor(new_point)
}

# Verify that Stream output matches Causal Batch output
all.equal(stream_output, causal_clean, tolerance = 1e-10)
```

## Comparison

Combining all three views illustrates the trade-off. The **Offline** method provides better smoothing (lower variance) because it "sees the future". The **Causal/Online** methods are reactive but naturally exhibit a slight phase lag.

```{r comparison}
ggplot(df_plot, aes(x = index)) +
  geom_line(aes(y = Offline, color = "Offline"), alpha = 0.8) +
  geom_line(aes(y = Causal, color = "Causal"), alpha = 0.8) +
  scale_color_manual(values = c("Offline" = "blue", "Causal" = "red")) +
  theme_minimal() +
  labs(title = "Trade-off: Smoothness vs. Causality") +
  coord_cartesian(xlim = c(500, 1000)) # Zoom in without dropping data
```
