---
title: "2. Benchmark: speed and reconstruction accuracy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Benchmark: speed and reconstruction accuracy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette compares `rLifting` against established R packages (`wavethresh`, `wavelets`, `adlift`, `waveslim`, `nlt`, `CNLTreg`) in terms of **execution speed** and **reconstruction accuracy (MSE)**.

We focus on the standard **offline denoising** task using the **Haar wavelet**.

## Data source

The results below are pre-computed using `microbenchmark` to ensure consistency.
The raw data is available in `data(benchmark_offline)`.

```{r setup}
library(rLifting)
library(ggplot2)
library(dplyr)

data("benchmark_offline", package = "rLifting")

head(benchmark_offline)
```

## 1. Execution speed

`rLifting` leverages a zero-allocation C++ core (via Rcpp), while many legacy packages rely on interpreted R code or older C interfaces.

```{r speed_plot}
# Calculate mean execution time
df_speed <- benchmark_offline %>%
  group_by(Pkg) %>%
  summarise(
    Mean_Time_s = mean(Time),
    Speedup = max(Mean_Time_s) / Mean_Time_s
  )

ggplot(df_speed, aes(x = reorder(Pkg, -Mean_Time_s), y = Mean_Time_s, fill = Pkg)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Execution Time (Lower is Better)",
    subtitle = "Denoising 1024 points (Haar, max levels) vs others",
    y = "Seconds",
    x = "Package"
  ) +
  theme_minimal()
```

The speed advantage of `rLifting` is often an order of magnitude, making it suitable for high-throughput pipelines.

## 2. Reconstruction accuracy (MSE)

Speed means nothing without accuracy. Here we compare the Mean Squared Error (MSE) of the denoised signal against the ground truth.

```{r mse_plot}
# Filter out packages where MSE wasn't calculated (e.g. pure transform benchmarks)
df_mse <- benchmark_offline %>%
  filter(!is.na(MSE))

ggplot(df_mse, aes(x = Pkg, y = MSE, fill = Pkg)) +
  geom_boxplot() +
  labs(
    title = "Reconstruction Error (MSE) - Lower is Better",
    subtitle = "Denoising Doppler Signal (SNR ~ 3dB)",
    y = "MSE"
  ) +
  theme_minimal()
```

`rLifting` achieves comparable or superior accuracy to established packages, validating the correctness of its Lifting Scheme implementation.
