---
title: "3. Causal and Stream Processing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Causal and Stream Processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

The introduction vignette presented offline, causal and stream as three modes with the same API. This vignette focuses on the two sliding-window modes: when to prefer them over offline, what they cost in MSE and latency, and how to set the parameters that they introduce. Causal and stream both add `window_size` and `update_freq` over the offline call; stream alone adds `irregular = TRUE` (at creation) and `t_val` (per sample on the closure), which is its mechanism for irregular-grid input. Causal accepts `t` as a single vector, the same way offline does. The empirical claims come from the regular-grid benchmark in `data/benchmark_rlifting.rda`; `window_size` and `update_freq` are kept fixed in that benchmark, so guidance on those axes is heuristic.

```{r setup}
library(rLifting)

if (!requireNamespace("ggplot2", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
  message("'ggplot2' is required to render plots. Vignette code will not run.")
} else {
  library(ggplot2)
}

data("doppler_example", package = "rLifting")
data("benchmark_rlifting", package = "rLifting")
set.seed(20260522)
```

## 1. Why causal mode exists

The offline path has full access to the signal: the output at sample $t$ can depend on every sample in the input. In many applications that assumption is invalid:

- **Backtesting financial models**: feeding tomorrow's prices into today's denoiser counts as look-ahead bias and inflates apparent performance.
- **Real-time control loops**: at the moment a controller fires, only past samples exist.
- **Streaming sensors**: data arrives one packet at a time; the denoised value must be emitted before the next packet is observed.

Causal mode replicates the streaming behaviour over a historical record (one batch call); stream mode is the same engine exposed as a sample-by-sample closure. Both rely on a fixed-size ring buffer holding the most recent `window_size` samples; nothing past sample $t$ is ever read.

## 2. The cost of causality

The bounded-history constraint has a measurable MSE cost. Below, the four DJ test signals denoised under the same wavelet (CDF 5/3), same boundary (`symmetric`), same shrinkage (semisoft, universal, default α/β), running offline against causal:

```{r causality-penalty}
sub = subset(
  benchmark_rlifting,
  Wavelet == "cdf53" & Boundary == "symmetric" &
    ThresholdMethod == "universal" & !grepl("tuned", Method) &
    Shrinkage == "semisoft"
)

penalty = data.frame(
  Signal = unique(sub$Signal),
  offline_MSE = sapply(unique(sub$Signal),
  function(s)
    sub$MSE_median[sub$Signal == s & sub$Mode == "offline"]),

  causal_MSE_settled = sapply(unique(sub$Signal),
  function(s)
    sub$MSE_settled_median[sub$Signal == s & sub$Mode == "causal"]),

  stream_MSE_settled = sapply(unique(sub$Signal),
  function(s)
    sub$MSE_settled_median[sub$Signal == s & sub$Mode == "stream"])
)

penalty$causal_over_offline = round(
  penalty$causal_MSE_settled / penalty$offline_MSE, 2
)

penalty
```

Two facts emerge from the table:

1. The causal penalty is signal-dependent: roughly 1.3× on `blocks` (discontinuities are easy to localise from past data) up to 3.3× on `heavisine` (smooth oscillations need broader support to estimate).
2. **Causal and stream MSEs are very close**, agreeing to about three decimal places in the table above. They share the same `WaveletEngine`, so for a fixed input signal under matching parameters the two paths agree to within numerical rounding; the small residual differences visible in the benchmark stem from the independent Monte Carlo noise realisations the benchmark assigns to each mode in parallel jobs. Choosing between them is an interface choice (batch call versus a per-sample closure), not an accuracy choice.

```{r causality-plot, echo=FALSE, fig.cap="Figure 1: Causality penalty across the four Donoho–Johnstone signals. Offline (left) uses the full signal; causal and stream (right) only past samples. Penalty ranges from 1.3× on blocks to 3.3× on heavisine."}
df_pen = data.frame(
  Signal = rep(penalty$Signal, 3),
  Mode = factor(
    rep(
      c("Offline", "Causal", "Stream"), 
      each = nrow(penalty)
    ),
    levels = c("Offline", "Causal", "Stream")
  ),
  
  MSE = c(
    penalty$offline_MSE,
    penalty$causal_MSE_settled,
    penalty$stream_MSE_settled
  )
)

ggplot(df_pen, aes(x = Signal, y = MSE, fill = Mode)) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c(
      "Offline" = "#0072B2", "Causal" = "#D55E00", "Stream" = "#009E73"
    )
  ) +
  theme_minimal() +
  labs(
    title = "Causality penalty: MSE by mode and signal",
    subtitle = "cdf53, symmetric, universal/semisoft, default α/β",
    x = NULL, y = "MSE (settled for causal/stream)"
  )
```

## 3. Choosing `window_size`

`window_size` ($W$) sets how many past samples the engine keeps in memory at any time. Larger $W$ gives a more stable MAD-based $\hat{\sigma}$ and finer spectral resolution; smaller $W$ gives lower memory footprint, shorter warm-up, and lower per-sample compute. The package's current benchmark fixed $W = 255$, so the recommendations below are heuristic — to be validated by a window-sweep extension to the benchmark.

```{r window-size-effect, eval=FALSE}
# Sketch of the trade-off (heuristic, not a chunk you should run blindly)
window_size = 63 # short — fast, lower resolution, ~31 raw samples at start
window_size = 127 # short-medium
window_size = 255 # benchmark default
window_size = 511 # long — better σ̂, more memory, longer warm-up
```

Guidelines that hold under the design assumptions in `inst/notes/02-adaptive-thresholding.md`:

- $W$ should be **at least** $16 \cdot 2^{\text{levels}}$ so the coarsest detail subband still holds a useful number of coefficients for the inverse transform.
- $W$ must be **odd** — the engine enforces this by adding 1 if an even value is supplied.
- Larger $W$ pushes the warm-up zone wider: the first $W - 1$ samples are returned as-is.

For most applications, $W \in [128, 512]$ balances stability and latency well. Lower values can be used if memory or warm-up cost dominates.

## 4. Choosing `update_freq`

`update_freq` is how often the adaptive threshold is recomputed. With `update_freq = 1` (default), the engine recomputes $\hat{\sigma}$ and the per-level $\lambda_k$ at every sample. With `update_freq = k`, it recomputes every $k$ samples and reuses the cached thresholds in between. The benchmark fixed `update_freq = 1`; the heuristic below comes from `inst/notes/02-adaptive-thresholding.md` §6.2:

| Noise behaviour | Recommended `update_freq` |
|:----------------|:---------------------------|
| Highly non-stationary (bursts, sudden regime shifts) | 1 |
| Mildly non-stationary | 5–20 |
| Stationary or quasi-stationary | 10–50 |
| Genuinely stationary, throughput-critical | 50–100 |

The CPU saving is roughly proportional to `update_freq`: the threshold computation is the dominant per-sample cost; once cached, the rest of the pipeline is fast. The benchmark does not currently quantify this trade-off; the per-sample numbers in §5 below all reflect `update_freq = 1`.

## 5. Per-sample latency

The streaming path is much slower than offline on a per-sample basis. Below, median per-sample time across all benchmark configurations, broken down by mode and wavelet:

```{r latency-table}
latency = aggregate(
  Per_sample_us_median ~ Mode + Wavelet, data = benchmark_rlifting,
  FUN = median
)

latency$Per_sample_us_median = round(latency$Per_sample_us_median, 2)

reshape(
  latency, idvar = "Wavelet", timevar = "Mode", direction = "wide"
)
```

```{r latency-plot, echo=FALSE, fig.cap="Figure 2: Median per-sample latency in microseconds across the three modes and five wavelets. Note the log scale — causal is ~80–115× slower per sample than offline, and stream adds another 1.5–2.1× over causal because of the R closure call overhead per sample."}
lat_long = aggregate(
  Per_sample_us_median ~ Mode + Wavelet, data = benchmark_rlifting,
  FUN = median
)

lat_long$Mode = factor(lat_long$Mode, levels = c("offline", "causal", "stream"))
ggplot(lat_long, aes(x = Wavelet, y = Per_sample_us_median, fill = Mode)) +
  geom_col(position = "dodge") +
  scale_y_log10() +
  scale_fill_manual(
    values = c(
      "offline" = "#0072B2", "causal" = "#D55E00", "stream" = "#009E73"
    )
  ) +
  theme_minimal() +
  labs(
    title = "Per-sample latency by mode and wavelet",
    subtitle = "Log scale on y; median across all benchmark configurations",
    x = NULL, y = "Microseconds per sample"
  )
```

Two practical takeaways:

- **Offline is 50–200× cheaper per sample** than the streaming modes. If the entire signal is available and you do not need leakage-free outputs, prefer offline.
- **Stream adds 50–110% on top of causal**. The extra cost is the R-side closure invocation per sample; the underlying C++ engine is the same. If you can batch the historical data, prefer `denoise_signal_causal` and pay the R overhead only once.

For a one-shot historical pass of N = 100,000 samples with `haar` at `update_freq = 1`, the table predicts roughly 0.6 s for causal and 1.3 s for stream.

## 6. Causal-specific wavelet recommendations

The wavelet ranking under causal/stream differs from the offline ranking because longer filters spend a higher fraction of the sliding window in the boundary zone and the per-window $\hat{\sigma}$ is more variable than the global offline estimate. Best wavelet per signal in causal mode (across all benchmark configurations):

```{r causal-best-wavelet}
best_causal = aggregate(
  MSE_settled_median ~ Signal + Wavelet,
  data = subset(benchmark_rlifting, Mode == "causal"),
  FUN = min
)

do.call(
  rbind,
  lapply(
    unique(best_causal$Signal),
    function(s) {
      ss = best_causal[best_causal$Signal == s, ]
      ss = ss[order(ss$MSE_settled_median), ]
      data.frame(
        Signal = s,
        ranking = paste(ss$Wavelet, collapse = " > ")
      )
    }
  )
)
```

Summary: `haar` is the dominant causal-mode wavelet on `blocks`, `bumps` and `heavisine`; `cdf97` retains its offline edge only on `doppler`. Use `haar` as the default for new causal/stream applications and switch to `cdf97` only when the signal is known to be smooth and oscillatory. The full mode-by-mode comparison is the subject of `vignette("07-benchmarks")`.

## 7. Leakage check by construction

Causality is a contract: the output at sample $t$ must not change if you alter samples after $t$. The package guarantees this through the ring buffer (no future read access), but it is healthy practice to verify it on your own signal. Below, a counterfactual leakage test:

```{r leakage-check}
x = doppler_example$noisy
scheme = lifting_scheme("haar")
ws = 255

base_out = denoise_signal_causal(
  x, scheme,
  window_size = ws, levels = 4,
  shrinkage = "semisoft"
)

# Counterfactual: corrupt the second half of the signal severely
x_perturbed = x
half = floor(length(x) / 2)
x_perturbed[(half + 1):length(x)] = x_perturbed[(half + 1):length(x)] +
  rnorm(length(x) - half, sd = 5)

perturbed_out = denoise_signal_causal(
  x_perturbed, scheme,
  window_size = ws, levels = 4,
  shrinkage = "semisoft"
)

# Outputs at positions ≤ half must be unchanged
max_diff_before = max(abs(base_out[1:half] - perturbed_out[1:half]))
max_diff_after  = max(
  abs(
    base_out[(half + 1):length(x)] -
      perturbed_out[(half + 1):length(x)]
  )
)

data.frame(
  region = c(
    "samples 1..half (before perturbation)",
    "samples (half+1)..n (after perturbation)"
  ),
  max_abs_diff = c(max_diff_before, max_diff_after)
)
```

The output on samples $1, \dots, \text{half}$ is bit-identical in the two runs (the corruption injected at $\text{half}+1$ cannot reach back). The output on samples $\text{half}+1, \dots, n$ differs, as expected. This is the leakage-free guarantee in action; the same test can be run on any user signal to confirm.

## 8. Decision guide

### 8.1 When the benchmark gives a direct answer

| Question | Recommendation | Evidence |
|:---------|:---------------|:---------|
| Which mode if I have the full signal and do not need leakage-free outputs? | `denoise_signal_offline()` | Per-sample latency 50–200× lower than causal/stream (§5 table) |
| Which mode if I have the full signal and need leakage-free outputs? | `denoise_signal_causal()` | Same engine and same MSE as stream, lower R overhead (§5 plot) |
| Which mode if I emit one sample at a time in production? | `new_wavelet_stream()` | Designed for the per-sample interface; stateful closure |
| Which wavelet for causal/stream? | `haar` by default; `cdf97` only if signal is smooth-oscillatory | causal best-wavelet table (§6) |
| Should I expect a quality cost vs offline? | Yes, 1.3×–3.3× MSE on the DJ benchmark | Causality penalty table (§2) |

### 8.2 Recommendations from the wavelet-denoising literature

The recommendations below come from established work in the wavelet-denoising literature and from `inst/notes/02-adaptive-thresholding.md`. They have empirical support in their original sources; the package's current benchmark, however, fixes `window_size = 255` and `update_freq = 1` and so has not independently re-confirmed them on rLifting's specific code paths. Treat them as well-established defaults whose translation to a particular workflow may benefit from a quick local check before being relied on.

| Situation | Recommendation |
|:----------|:---------------|
| Memory/latency budget is the bottleneck | smaller `window_size` (63–127); accept a noisier $\hat{\sigma}$ |
| Quality is the bottleneck, memory is not | larger `window_size` (511+); longer warm-up but more stable thresholds |
| Stationary noise, want lower CPU | `update_freq = 10–50` |
| Highly non-stationary noise (bursts) | `update_freq = 1` |
| Causal denoising on irregular grids | `denoise_signal_causal(x, scheme, t = t_phys, ...)` accepts `t` the same way as offline |

For the broader empirical context — mode-by-mode wavelet comparisons, regular- and irregular-grid results, and package-against-package benchmarks — see `vignette("07-benchmarks")` and the practical guidance in `vignette("01-introduction")` §3. For the threshold subsystem covered briefly in §4 here, see `vignette("02-thresholding-and-tuning")`. The operational reference for the adaptive threshold (formulas for $\hat{\sigma}$, $\lambda_k$, the four shrinkages, and the engine's update protocol) is `inst/notes/02-adaptive-thresholding.md`.

## References

Donoho, D. L., & Johnstone, I. M. (1994). Ideal spatial adaptation by wavelet shrinkage. *Biometrika*, **81**(3), 425–455.

Liu, Z., Mi, Y., & Mao, Y. (2014). Improved real-time denoising method based on lifting wavelet transform. *Measurement Science Review*, **14**(3), 152–159. DOI: 10.2478/msr-2014-0020.

Sweldens, W. (1996). The lifting scheme: A custom-design construction of biorthogonal wavelets. *Applied and Computational Harmonic Analysis*, **3**(2), 186–200.
