---
title: "3. Causal decomposition: the rlifting advantage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Causal decomposition: the rlifting advantage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Most wavelet packages in R are designed for **offline analysis**: they assume the entire signal is available at once. To perform "causal" filtering (where time $t$ depends only on $0 \dots t$) with these packages, one must implement a **sliding window** loop, re-calculating the transform for every new point.

This approach is $O(N \cdot W)$ (where $W$ is the window size).

`rLifting`, in contrast, implements a specialized **causal mode** using a ring buffer architecture, which updates the transform state in $O(N)$ (amortized $O(1)$ per point).

## 1. Speed comparison: optimized vs. naive

We compare `rLifting::denoise_signal_causal` against a naive sliding-window implementation using `wavethresh`.

```{r causal_speed}
data("benchmark_causal", package = "rLifting")

# Data is a list
time_rl <- benchmark_causal$rLifting_Time_Avg
time_naive <- benchmark_causal$Wavethresh_Naive_Time
speedup <- benchmark_causal$Speedup_Factor

df_causal <- data.frame(
  Method = c("rLifting (Optimized)", "Naive Loop (Others)"),
  Time = c(time_rl, time_naive)
)

knitr::kable(df_causal, caption = "Execution Time for Causal Processing")
```

**Speedup Factor:** `r round(speedup, 1)`x.

The massive difference confirms that using standard packages for real-time sliding window analysis is computationally infeasible for high-frequency data.

## 2. Zero-leakage (impulse response)

A true causal filter must have **zero lookahead**. This means an impulse at time $t$ must NOT affect the output at any time $t' < t$.

We verify this by analyzing the **impulse response**.

```{r leakage}
library(ggplot2)
data("leakage_results", package = "rLifting")

knitr::kable(leakage_results, caption = "Pre-Trigger Energy (Data Leakage)")
```

- **Offline mode:** The energy leaks into the past (Pre-Trigger Energy > 0). The filter "sees" the impulse coming and reacts before it happens.
- **Causal mode:** The Pre-Trigger Energy is exactly 0. The filter only reacts *after* or *at* the moment the impulse occurs.

This property is critical for financial backtesting (avoiding look-ahead bias) and real-time control systems.
