---
title: "4. Real-time signal smoothing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Real-time signal smoothing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates the **real-time** capabilities of `rLifting`.
We simulate a high-frequency data stream (e.g., from a sensor) and process it point-by-point.

## Simulation setup

We generate a noisy sine wave to simulate a continuous signal.

```{r setup}
library(rLifting)
library(ggplot2)

# Simulate 1000 data points
set.seed(42)
n <- 1000
t <- seq(0, 10*pi, length.out = n)
clean <- sin(t)
noisy <- clean + rnorm(n, sd = 0.5)

# Initialize the stream processor
# Uses a sliding window of 128 points
# Algorithm: Haar wavelet (Level 3)
stream_proc <- new_wavelet_stream(
  scheme = lifting_scheme("haar"),
  window_size = 128,
  levels = floor(log2(128)),
  method = "soft",
  alpha = 0.5, # Slightly more aggressive smoothing
  beta = 1.0
)
```

## Processing loop

In a real application, this loop would be replaced by an event listener (WebSocket, MQTT, Serial Port).
Here, we iterate over the vector.

```{r loop}
output <- numeric(n)
process_time <- numeric(n)

for (i in 1:n) {
  # Measure time per sample
  start <- Sys.time()
  
  # Feed one sample, get one denoised sample
  output[i] <- stream_proc(noisy[i])
  
  end <- Sys.time()
  process_time[i] <- as.numeric(end - start)
}

# The first 'window_size' samples are returned raw (warm-up)
# Let's inspect the results
df <- data.frame(
  Time = 1:n,
  Noisy = noisy,
  Clean = output,
  Truth = clean
)
```

## Latency analysis

Because the core engine is written in zero-allocation C++, the per-sample overhead is extremely low.

```{r latency}
# Average latency in microseconds
avg_latency_us <- mean(process_time) * 1e6
cat(sprintf("Average Latency per Sample: %.2f microseconds\n", avg_latency_us))

# Plot latency distribution (excluding first compilation hit)
hist(process_time[-1] * 1e6, 
     main = "Processing Latency per Point", 
     xlab = "Microseconds", 
     col = "steelblue", 
     breaks = 30)
```

## Visualizing the result

```{r plot}
# Visualize a segment after warm-up
ggplot(subset(df, Time > 200 & Time < 600), aes(x = Time)) +
  geom_line(aes(y = Noisy), color = "grey80") +
  geom_line(aes(y = Truth, color = "Truth"), linetype = "dashed") +
  geom_line(aes(y = Clean, color = "Real-Time Filter"), linewidth = 1) +
  scale_color_manual(values = c("Truth" = "black", "Real-Time Filter" = "red")) +
  theme_minimal() +
  labs(title = "Real-Time Smoothing of Noisy Sine Wave")
```

The filter successfully recovers the sine wave shape from the noise, with a minimal phase delay inherent to causal processing.
