---
title: "5. Irregular Grids"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Irregular Grids}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

Most wavelet denoising literature assumes the signal arrives on a uniform grid. Many real signals do not: ECG with variable RR intervals, IoT streams with dropped packets, financial tick data, geophysical sensors. `rLifting` handles these natively by carrying the physical sample positions `t` through the predict step and replacing the fixed-coefficient predictor with Lagrange interpolation at the actual time of the missing sample. This vignette gives the mechanism in §2–§3, the API in §4, and then a *single robust default per mode* (§5) chosen from the benchmark in `data(benchmark_rlifting_irregular)` to be never worse than ~2× from the per-signal optimum, with every component mechanically justifiable. §6–§8 explain why it works; §9 gives narrow recommendations for when to deviate.

```{r setup}
library(rLifting)

if (!requireNamespace("ggplot2", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
  message("'ggplot2' is required to render plots. Vignette code will not run.")
} else {
  library(ggplot2)
}

data("benchmark_rlifting_irregular", package = "rLifting")
set.seed(20260605)
```

## 1. What makes a grid irregular

A grid is irregular when the spacings $\Delta_i = t_{i+1} - t_i$ are not constant. Two questions decide what to do with it:

- Is the irregularity informational or incidental? Tick data over-sampling high-activity periods carries information in the spacing; dropped packets on an otherwise periodic sensor do not.
- How extreme is the spacing variation? Measured by the coefficient of variation $\mathrm{CV}(\Delta) = \mathrm{sd}(\Delta) / \mathrm{mean}(\Delta)$. Low CV is "almost uniform"; high CV breaks the uniform-grid approximation outright.

The benchmark covers seven test signals (three physically motivated — `linear_phys`, `trend_events`, `blocks_gapped` — and four Donoho–Johnstone classics re-sampled on non-uniform grids); it does not parametrise spacing CV.

## 2. Mechanism — Lagrange in the predict step

On a uniform grid the predict step of an interpolating wavelet assumes each odd sample sits at a known fractional position between its even neighbours. On an irregular grid that assumption fails. When `t` is supplied, `rLifting` replaces the fixed-coefficient predict with a Lagrange polynomial of degree $k - 1$ (where $k$ is the predict filter length) evaluated at the actual position of the odd sample. The update step is unchanged: it acts on the approximation subband and does not depend on positions.

The same mechanism runs in every mode. The only piece that touches user code is how `t` is supplied (§4). The implementation walk-through, including the linear-extrapolation rule for positions at the boundaries, is in `inst/notes/01-lifting-scheme-and-transform.md` §8.

## 3. Wavelet eligibility

A predict step is "interpolating" — eligible for the Lagrange path — when its coefficients sum to one (partition of unity). `lifting_scheme()` checks this condition and stores the polynomial degree in the `degree` field of each step.

| Wavelet | `degree` | Behaviour when `t` is supplied |
|:--------|:--------:|:-------------------------------|
| `haar`   | 0 | nearest-neighbour (Haar requires no interpolation; position-aware ≡ position-ignoring) |
| `cdf53`  | 1 | linear interpolation at the actual odd position |
| `dd4`    | 3 | cubic Lagrange interpolation |
| `db2`    | $-1$ | predict has $\sum c_k \neq 1$; `t` is ignored, warning emitted |
| `cdf97`  | $-1$ | same as `db2` |
| `lazy`   | n/a  | `t` is ignored (no predict step requires interpolation) |

When the warning fires the wavelet still runs and produces a valid denoised signal — just without Lagrange correction. The output is identical to the regular-grid call.

## 4. The API in each mode

```{r api, eval = FALSE}
# Offline and causal accept t as a vector parallel to the signal.
denoise_signal_offline(x, scheme, t = t_phys)
denoise_signal_causal(x, scheme, t = t_phys, window_size = 255)

# Stream creates the processor with irregular = TRUE,
# then accepts t_val per sample.
stream = new_wavelet_stream(scheme, irregular = TRUE, window_size = 255)
for (i in seq_along(x)) y[i] = stream(x[i], t_val = t_phys[i])

# Step-by-step lwt() also accepts t; the returned `lwt` object
# carries it forward to ilwt() automatically.
lwt_obj = lwt(x, scheme, t = t_phys)
x_back  = ilwt(lwt_obj)
```

`t` must be strictly increasing. The R wrappers enforce sortedness in batch calls; for the stream closure, monotonic delivery is the caller's responsibility.

## 5. Two robust defaults

For offline mode:

```{r}
denoise_signal_offline(
  x, lifting_scheme("cdf53"), t = t_phys,
  extension = "local_linear",
  threshold_method = "sure",
  shrinkage = "scad"
)
```

For causal and stream:

```{r}
tuned = tune_alpha_beta(x_slice, lifting_scheme("haar"), levels = 4)

denoise_signal_causal(
  x, lifting_scheme("haar"), t = t_phys,
  window_size = 255,
  extension = "zero",
  threshold_method = "universal",
  shrinkage = "scad",
  alpha = tuned$alpha,
  beta = tuned$beta
)
```

These defaults were chosen to be (a) never far from the per-signal best across the benchmark, and (b) mechanically justifiable in every component. §6 explains the per-component choices; §7 quantifies how close to optimal each default is.

## 6. Why these choices

### Offline: cdf53 + local_linear + SURE + SCAD

- `cdf53` is the *simplest* interpolating wavelet ($\mathrm{degree} = 1$, two-tap predict). Linear interpolation handles smooth regions exactly and degrades gracefully near transitions. Higher-order wavelets (`dd4`) match smooth signals better but pay a larger boundary contamination zone per level (`vignette("04-boundary-modes")` §1).
- `extension = "local_linear"` fits an OLS line through the `ll_k` nearest boundary samples and evaluates it at the virtual extrapolation index. On smooth-trend signals this preserves the local linear behaviour into the virtual region. The default `symmetric` (half-sample reflection) instead pins the virtual value to the boundary sample, which on a trending signal injects a phantom *flattening* of the trend — small enough to survive thresholding, large enough to bias the approximation subband.
- `threshold_method = "sure"` estimates a per-level $\hat\sigma_k$ and minimises the per-level Stein risk to pick the threshold. On signals with non-uniform spacing the detail-coefficient distribution at each level depends on the local spacing pattern, so the SURE per-level estimate beats the global MAD-plus-recursive-decay of the universal rule.
- `shrinkage = "scad"` is identity for $|d| > a\lambda$ (Antoniadis & Fan, 2001), so large coefficients pass through unbiased — critical when the signal has isolated jumps. The continuous transition at $\lambda$ avoids the Gibbs-style ringing of hard.

### Causal / stream: haar + zero + tuned universal + SCAD

- `haar` has a single-tap predict (`degree = 0`, nearest-neighbour), so the boundary contamination zone at every level is one sample — the smallest possible. Note that for `haar`, position-aware ≡ position-ignoring: its nearest-neighbour predict works correctly on irregular grids without position correction — the local difference `d[i] = x[2i+1] - x[2i]` is valid regardless of spacing, so the output is identical with or without `t`. In a sliding window of `window_size = 255`, the per-sample work is dominated by threshold update on the finest subband (`vignette("03-causal-stream")` §3). A short filter is the right call when the per-window estimate is already noisy.
- `extension = "zero"` pads missing boundary taps with zero — a known, fixed value. Benchmark data confirms that for `haar` in causal mode, `zero` and `one_sided` produce similar MSE on all seven irregular-grid signals (the single-tap predict collapses the two modes). `zero` is preferred over `one_sided` here because it does not skip the Lagrange path: if the user later switches to an interpolating wavelet (`cdf53`, `dd4`), the `t` positions will be respected without any code change.
- `tune_alpha_beta()` + `universal` + `scad`: in causal mode the per-window MAD already adapts $\hat\sigma$ to the local noise; the universal rule with tuned $(\alpha, \beta)$ closes the remaining gap. SURE in causal mode has too few finest-level coefficients per window ($m_1 = W/2 \approx 127$) to be reliable.

## 7. How close to optimal these defaults are

For each signal, compare the robust default against the per-signal best `(wavelet, boundary, method)` triple:

```{r off-penalty}
df = benchmark_rlifting_irregular
off = df[df$Mode == "offline" & !is.na(df$MSEpos_median), ]

robust_off = off[
  off$Wavelet == "cdf53" & 
    off$Boundary == "local_linear" &
    off$Method == "sure_scad",
  c("Signal", "MSEpos_median")
]
names(robust_off)[2] = "robust_MSE"

best_off = do.call(
  rbind, 
  by(
    off, off$Signal, function(s) {
      i = which.min(s$MSEpos_median)
      data.frame(
        Signal = s$Signal[i],
        best_MSE = s$MSEpos_median[i],
        best_config = paste(
          s$Wavelet[i], s$Boundary[i],
          s$Method[i], sep = "/"
        )
      )
    }
  )
)

cmp_off = merge(robust_off, best_off, by = "Signal")
cmp_off$penalty = round(cmp_off$robust_MSE / cmp_off$best_MSE, 2)
cmp_off$robust_MSE = round(cmp_off$robust_MSE, 4)
cmp_off$best_MSE = round(cmp_off$best_MSE, 4)
knitr::kable(
  cmp_off[
    , 
    c(
      "Signal", "robust_MSE", "best_MSE",
      "penalty", "best_config"
    )
  ],
  row.names = FALSE,
  caption = "Offline robust default (cdf53 / local_linear / sure_scad) vs per-signal best."
)
```

The robust default never pays more than ~2.4× over the per-signal optimum, and is within 1.5× on five of seven signals. The 2× ceiling on Blocks-type signals (`blocks_dj_irr`, `blocks_gapped`) is the cost of using a linear-interpolation predict on a piecewise-constant signal — see §9 for when to deviate.

For causal mode:

```{r cs-penalty}
cs = df[df$Mode == "causal" & !is.na(df$MSEpos_median), ]
robust_cs = cs[
  cs$Wavelet == "haar" & cs$Boundary == "zero" &
    cs$Method == "universal_tuned_scad",
  c("Signal", "MSEpos_median")
]

names(robust_cs)[2] = "robust_MSE"

best_cs = do.call(
  rbind, 
  by(
    cs, cs$Signal, function(s) {
      i = which.min(s$MSEpos_median)
      data.frame(
        Signal = s$Signal[i],
        best_MSE = s$MSEpos_median[i],
        best_config = paste(s$Wavelet[i], s$Boundary[i], s$Method[i], sep = "/")
      )
    }
  )
)

cmp_cs = merge(robust_cs, best_cs, by = "Signal")
cmp_cs$penalty = round(cmp_cs$robust_MSE / cmp_cs$best_MSE, 2)
cmp_cs$robust_MSE = round(cmp_cs$robust_MSE, 4)
cmp_cs$best_MSE = round(cmp_cs$best_MSE, 4)
knitr::kable(
  cmp_cs[
    , 
    c(
      "Signal", "robust_MSE", "best_MSE",
      "penalty", "best_config"
    )
  ],
  row.names = FALSE,
  caption = "Causal robust default (haar / zero / universal_tuned_scad) vs per-signal best."
)
```

The causal default is within 1.15× of optimum on every signal. The choice of `haar + zero` is so consistently good in causal mode that there is rarely a reason to deviate. Note that for `haar`, `zero` and `one_sided` produce identical MSE (the nearest-neighbour predict collapses both boundary modes to the same computation); `zero` is preferred here because it leaves the Lagrange path open for users who switch to an interpolating wavelet.

## 8. Does position-aware processing actually help?

With the robust default in place, compare `MSEpos` (irregular path active) against `MSEfix` (uniform-grid baseline) for the same configuration:

```{r pos-fix}
robust_off_pf = off[
  off$Wavelet == "cdf53" & 
    off$Boundary == "local_linear" &
    off$Method == "sure_scad",
  c("Signal", "MSEpos_median", "MSEfix_median")
]

robust_off_pf$ratio = round(
  robust_off_pf$MSEpos_median / robust_off_pf$MSEfix_median, 2
)

robust_off_pf$MSEpos = round(robust_off_pf$MSEpos_median, 4)
robust_off_pf$MSEfix = round(robust_off_pf$MSEfix_median, 4)

knitr::kable(
  robust_off_pf[, c("Signal", "MSEpos", "MSEfix", "ratio")],
  row.names = FALSE,
  caption = "Robust default: position-aware vs uniform-grid baseline, offline mode."
)
```

Two regimes appear:

- On the DJ-irregular signals and `blocks_gapped`, the ratio is ~1: position-aware and uniform-grid give essentially the same result. The benefit of carrying `t` is small or absent.
- On `linear_phys` and `trend_events`, position-aware is 50–100× better than uniform-grid (`linear_phys`: ~95×; `trend_events`: ~72×): enabling the Lagrange path is unambiguously the right choice.

The mechanism is symmetric to what §6 said about the boundary choice. On smooth-trending signals the uniform-grid approximation forces the predict step to assume a fixed midpoint between even neighbours — wrong by a factor proportional to the local spacing variation. The Lagrange-corrected predict matches the actual midpoint of the smooth trend and removes the bias.

Practical implication: passing `t` to the robust default is never harmful and is critical for trend-like signals. This is the simple version of the question the regular-grid vignettes leave hanging.

## 9. When to deviate from the robust default

| Situation | Change | Why |
|:----------|:-------|:----|
| You know the signal is mostly piecewise-constant (Blocks-like) | switch `cdf53` → `haar`, keep `local_linear` and `universal_hard` instead of `sure_scad` | linear interpolation injects ringing across true discontinuities; haar's single-tap predict and hard thresholding preserve jumps exactly |
| Causal/stream and high spacing CV (>50%) | keep the default | the empirical winner across all seven signals; no diagnostic that triggers a switch |
| Offline and you have time for a per-signal sweep | run `tune_alpha_beta()` and try `(symmetric, periodic, local_linear)` × `(universal_tuned_scad, sure_scad)` | the §7 table shows the achievable optimum is ~2× below the robust default's MSE on most signals; that gap is closeable with a small grid search |

### 9.1 Pitfalls

- Unsorted `t`: hard error in `denoise_signal_offline`, `denoise_signal_causal`, and `lwt`. The stream closure does not validate per call — caller's responsibility.
- `db2` or `cdf97` with `t`: warning emitted; `t` is ignored and the wavelet runs as if on a uniform grid. If irregular handling is required, switch to `cdf53` or `dd4`.
- `extension = "one_sided"` with `t`: warning emitted; the engine takes the `one_sided` branch before the irregular branch, skipping Lagrange. In offline mode use `local_linear` instead. In causal mode the robust default uses `zero`, which is empirically equivalent to `one_sided` for `haar` (identical MSE on all benchmark signals) and does not skip Lagrange if the user switches to an interpolating wavelet.
- Position extrapolation at the boundaries uses unconditional linear extrapolation of `t`, independently of the `extension` parameter (which controls *values*, not positions). See `inst/notes/04-boundary-and-threshold.md` §A.4.

## 10. Worked example — Doppler on a log-normal grid

```{r worked-example, fig.cap = "Top: noisy Doppler signal on a log-normal-spaced grid (CV approx 0.3). Bottom: offline denoising with the §5 robust default versus the uniform-grid baseline (t ignored).", fig.height = 6}
n = 512
dt = exp(rnorm(n, sd = 0.3))
dt = dt / mean(dt)
t_phys = cumsum(dt)
t_phys = t_phys / max(t_phys)

doppler = function(t) sqrt(t * (1 - t)) * sin(2 * pi * 1.05 / (t + 0.05))
clean = doppler(t_phys)
noisy = clean + rnorm(n, sd = 0.15)

scheme = lifting_scheme("cdf53")

# Baseline: ignore t.
y_fix = denoise_signal_offline(
  noisy, scheme, levels = 4,
  threshold_method = "sure", shrinkage = "scad",
  extension = "local_linear"
)

# Robust default: pass t = t_phys.
y_pos = denoise_signal_offline(
  noisy, scheme, t = t_phys, levels = 4,
  threshold_method = "sure", shrinkage = "scad",
  extension = "local_linear"
)

dat = data.frame(
  t = rep(t_phys, 3),
  y = c(noisy, y_fix, y_pos),
  series = factor(
    rep(
      c(
        "noisy", "uniform-grid baseline (t ignored)",
        "robust default with t = t_phys"
      ), 
      each = n
    ),
    levels = c(
      "noisy", "uniform-grid baseline (t ignored)",
      "robust default with t = t_phys"
    )
  )
)

ggplot(dat, aes(x = t, y = y)) +
  geom_line(linewidth = 0.4) +
  facet_wrap(~ series, ncol = 1, scales = "free_y") +
  labs(x = "physical time t", y = NULL) +
  theme_minimal(base_size = 10)

mse = function(a, b) mean((a - b) ^ 2)
mses = data.frame(
  strategy = c(
    "uniform-grid baseline (t ignored)",
    "robust default with t = t_phys"
  ),
  MSE = round(c(mse(clean, y_fix), mse(clean, y_pos)), 4)
)

knitr::kable(
  mses, row.names = FALSE,
  caption = "MSE against the clean signal on this realisation."
)
```

On this realisation the two strategies are close: Doppler is one of the DJ-irregular signals where §8 reported the position-aware / uniform-grid ratio is essentially 1. The structural advantage of passing `t` would be more visible on a `linear_phys`- or `trend_events`-style signal — change `doppler(t)` to `t` and the gap widens substantially.

## 11. Where to look next

- For the implementation of Lagrange in the predict step and the per-level position tracking, see `inst/notes/01-lifting-scheme-and-transform.md` §8.
- For boundary handling of values vs positions, see `inst/notes/04-boundary-and-threshold.md` §A.4.
- For the cross-package benchmark comparing rLifting irregular against `adlift` and `nlt`, see `vignette("07-benchmarks")`; the raw data already ships in `data(benchmark_adlift_irregular)`, `data(benchmark_nlt_irregular)`, and `data(benchmark_rlifting_irregular)`.

## References

Antoniadis, A., & Fan, J. (2001). Regularization of wavelet approximations. *Journal of the American Statistical Association*, **96**(455), 939–967.

Daubechies, I., Guskov, I., Schröder, P., & Sweldens, W. (1999). Wavelets on irregular point sets. *Philosophical Transactions of the Royal Society A*, **357**(1760), 2397–2413.

Donoho, D. L., & Johnstone, I. M. (1995). Adapting to unknown smoothness via wavelet shrinkage. *Journal of the American Statistical Association*, **90**(432), 1200–1224.

Sweldens, W. (1996). The lifting scheme: a custom-design construction of biorthogonal wavelets. *Applied and Computational Harmonic Analysis*, **3**(2), 186–200.
