---
title: "5. Extensions: custom wavelets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Extensions: custom wavelets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

One of the strengths of the Lifting Scheme is its modularity. Unlike standard convolution-based wavelets, Lifting allows you to construct wavelets by composing simple **predict (P)** and **update (U)** steps.

`rLifting` allows advanced users to define custom wavelets.

## Defining a custom wavelet

A wavelet is defined by a sequence of steps and a normalization factor.
Let's recreate the **linear (Bernstein)** wavelet manually.

```{r custom}
library(rLifting)

# Step 1: Predict (P)
# Predicts odd samples from even samples using linear interpolation
# coeffs = c(-0.5, -0.5) implies: d[i] -= 0.5*s[i] + 0.5*s[i+1]
p_step <- lift_step(
  type = "predict", 
  coeffs = c(-0.5, -0.5), 
  position = "center"
)

# Step 2: Update (U)
# Updates even samples to preserve the average
# coeffs = c(0.25, 0.25) implies: s[i] += 0.25*d[i-1] + 0.25*d[i]
u_step <- lift_step(
  type = "update", 
  coeffs = c(0.25, 0.25), 
  position = "center"
)

# Construct the Scheme
my_linear_wavelet <- custom_wavelet(
  name = "LinearManual",
  steps = list(p_step, u_step),
  norm = c(sqrt(2), 1/sqrt(2)) # Standard biorthogonal normalization
)

print(my_linear_wavelet)
```

## Using the custom wavelet

Once defined, this custom wavelet can be used in any `rLifting` function (offline or online).

```{r usage}
# Test signal
x <- 1:16

# Forward Transform
fwd <- lwt(x, my_linear_wavelet, levels = floor(log2(length(x))))
print(fwd$coeffs$d1) # Should be zero for a linear signal!

# Inverse Transform
rec <- ilwt(fwd)
all.equal(x, rec)
```

## Diagnosing the custom wavelet

You can use the `diagnose_wavelet` tool to verify the properties of your new creation.

```{r diagnosis}
# Check for Vanishing Moments (Order 2 = Linear)
diagnose_wavelet(
  my_linear_wavelet, 
  config = list(is_ortho = FALSE, vm_degrees = c(0, 1), max_taps = 5),
  plot = FALSE
)
```
