---
title: "7. Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{7. Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

This vignette reports the empirical comparison of `rLifting` against three established R packages — `wavethresh` (decimated DWT, regular grids), `adlift` (adaptive lifting on irregular grids), and `nlt` (nondecimated lifting on irregular grids) — and the internal comparison of rLifting's three modes (offline, causal, stream). The aim is to give an honest reading of where rLifting sits in the design space: which signals it wins on, which it loses, the speed–quality trade-offs, and the regimes where each package is the right choice.

All numbers come from the bundled benchmark datasets and are reproducible from the code chunks below.

```{r setup}
library(rLifting)
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
  message("'ggplot2' is required to render plots. Vignette code will not run.")
} else {
  library(ggplot2)
}

data("benchmark_rlifting", package = "rLifting")
data("benchmark_wavethresh", package = "rLifting")
data("benchmark_adlift", package = "rLifting")
data("benchmark_nlt", package = "rLifting")
data("benchmark_rlifting_irregular", package = "rLifting")
data("benchmark_adlift_irregular", package = "rLifting")
data("benchmark_nlt_irregular", package = "rLifting")
```

## 1. Methodology and scope

Each benchmark cell is a `(signal, package, configuration)` triple evaluated over $N_\text{sim} = 1000$ Monte Carlo replications, each replication being an independent draw of Gaussian noise added to a deterministic signal. For each cell we report seven order statistics — min, q1, median, mean, q3, max, se — for both reconstruction MSE (against the noise-free signal) and wall time. The "best" of a configuration set means the one with the lowest *median* MSE.

Regular grid. Four Donoho–Johnstone test signals (`blocks`, `bumps`, `doppler`, `heavisine`), each of length $N = 1024$, with i.i.d. Gaussian noise at $\sigma = 0.3$. The same signals and the same noise distribution apply to all four packages, so the regular-grid MSE comparison is apples-to-apples on the signal and noise dimensions. Each package was swept over its own parameter space (rLifting: 6 wavelets × 5 boundaries × 12 thresholding methods × 3 modes; wavethresh: 8 wavelets × 22 boundary/threshold/shrinkage combinations; adlift and nlt: 4 predictor families × 24 option combinations); the "best" therefore picks the highest-quality reachable configuration in each package on its own terms.

Irregular grid. Seven test signals: the four Donoho–Johnstone classics resampled at non-uniform positions (`blocks_dj_irr`, `bumps_dj_irr`, `doppler_dj_irr`, `heavisine_dj_irr`) and three physically motivated signals (`linear_phys`, `trend_events`, `blocks_gapped`). Per-signal noise standard deviations range from 0.15 to 0.50 depending on the signal scale (see `?benchmark_rlifting_irregular`). All three irregular-grid packages saw the same realisations.

### 1.1 Which package appears where, and why

The two grids cover different design surfaces, and the package selection reflects each package's native regime:

- `wavethresh` appears only in §2–§3 (regular grid). It implements the decimated DWT and assumes uniform sampling. It can be applied to irregularly-sampled data only by first resampling to a uniform grid — but that introduces a *choice of resampling method* (linear interpolation, spline, kernel regression, ...) that would dominate the empirical comparison and hide which package's denoising primitive actually does the work. The cleanest read of wavethresh's capability is on its native regime; the irregular comparison therefore restricts to packages that handle non-uniform sampling natively.

- `rLifting` appears in both §2 and §5. Including it on the regular grid (§2) answers a fair question: *is rLifting's native irregular-grid support free, or does the engine pay a quality penalty when used on the regime where `wavethresh` is the established reference?* The §2 numbers settle it — on the four DJ signals rLifting trails wavethresh by ~18% at the geometric mean MSE level while running ~40× faster per sample. Including it on the irregular grid (§5) puts it head-to-head with the two native irregular packages (`adlift`, `nlt`).

- `adlift` and `nlt` appear in both §2 and §5. They are both irregular-grid lifting packages but also ship regular-grid implementations. Including them in §2 lets us check whether the irregular machinery costs anything in MSE when fed a uniform grid (it doesn't: adlift sits within 12% of wavethresh at the geomean level). Including them in §5 is the main cross-package comparison on the irregular grid.

- rLifting's causal and stream modes (§4 and §7) have no counterpart in the other three packages. None of `wavethresh`, `adlift`, or `nlt` implements sliding-window denoising in either grid. The §4/§7 comparisons are therefore intra-rLifting only — but the modes are themselves part of what rLifting offers and the speed–quality trade-off they expose is unique to this package in the comparison set.

### 1.2 What is *not* measured

Non-Gaussian noise, non-stationary noise, signal lengths other than $N = 1024$, real-world signals (ECG, financial, geophysical), pathologies of the irregular grid beyond the seven canonical cases, and the `update_freq` knob in causal/stream modes (held at 1 throughout). The conclusions below should not be extrapolated past these boundaries without a local check.

## 2. Regular grid — MSE

### Best achievable configuration

The question this section answers: *if you have uniformly-sampled data, is there a quality cost to using `rLifting` instead of the established regular-grid reference (`wavethresh`)?* For each (package, signal) pair we report the median MSE at the package's empirically best configuration:

```{r reg-mse}
reg_best = function(df, mse_col, time_col, pkg_label) {
  do.call(
    rbind, 
    by(
      df, df$Signal, function(s) {
        i = which.min(s[[mse_col]])
        data.frame(
          Signal = s$Signal[i],
          Package = pkg_label,
          MSE = s[[mse_col]][i],
          us = s[[time_col]][i] * 1e6 / s$N[i]
        )
      }
    )
  )
}

off_reg = benchmark_rlifting[benchmark_rlifting$Mode == "offline", ]
reg = rbind(
  reg_best(off_reg, "MSE_median", "Time_total_median", "rLifting (offline)"),
  reg_best(benchmark_wavethresh, "MSE_median", "Time_median", "wavethresh"),
  reg_best(benchmark_adlift, "MSE_median", "Time_median", "adlift"),
  reg_best(benchmark_nlt, "MSE_median", "Time_median", "nlt")
)

reg_wide = reshape(
  reg[, c("Signal", "Package", "MSE")],
  idvar = "Signal", timevar = "Package", direction = "wide"
)
names(reg_wide) = sub("MSE\\.", "", names(reg_wide))

knitr::kable(
  reg_wide, row.names = FALSE, digits = 4,
  caption = "Regular-grid MSE at each package's best configuration."
)
```

```{r reg-mse-fig, fig.cap = "Regular-grid MSE at each package's best configuration, by signal."}
ggplot(reg, aes(x = Package, y = MSE, fill = Package)) +
  geom_col() +
  facet_wrap(~ Signal, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Median MSE") +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

Reading the table: `wavethresh` is the clear leader, winning on three of four signals; `rLifting` wins only on `bumps`. The gap from `rLifting` to the winner ranges from 0% (`bumps`) to ~39% (`blocks`) and ~29% (`doppler`). `adlift` is competitive on the smooth signals (`doppler`, `heavisine`) but trails on `blocks` and `bumps`; `nlt` is consistently last. §2.2 and §3 will show what wavethresh's MSE lead costs: it requires the most tuning to reach its best (§2.3) and runs ~40× slower per sample (§3).

As a single-number summary, the geometric mean MSE across the four signals:

```{r reg-mse-geom}
geom_mean = function(x) exp(mean(log(x)))
geom_tbl = aggregate(MSE ~ Package, data = reg, FUN = geom_mean)
geom_tbl$MSE = round(geom_tbl$MSE, 4)
geom_tbl = geom_tbl[order(geom_tbl$MSE), ]
knitr::kable(
  geom_tbl, row.names = FALSE,
  caption = "Geometric mean MSE across the four DJ signals."
)
```

`wavethresh` leads at the geomean level (0.0096), followed by `adlift` (0.0108, ~12% above wavethresh), `rLifting` (0.0113, ~18% above wavethresh), and `nlt` (0.0150, ~33% above rLifting). rLifting pays a modest but real ~18% geomean penalty relative to wavethresh — a package built specifically for this regime. It wins outright on `bumps` and is the fastest package by far (§3); whether the geomean gap matters depends on the application.

### Out-of-the-box configuration

The best-configuration comparison above requires knowing which wavelet, boundary, and threshold method to use — knowledge that comes from reading the vignettes or running a sweep. The table below fixes a single default for each package and measures what you get on the first call without tuning:

- rLifting: `haar` + `symmetric` + universal threshold + semisoft shrinkage (offline).
- wavethresh: filter `db1` (Haar) + `BayesThresh` policy (soft) — the package's default threshold method.
- adlift / nlt: `LinearPred` predictor, 1 neighbour, interpolating, no closure, mean aggregation — the simplest fixed predictor in each package.

```{r reg-default}
rl_def = benchmark_rlifting[
  benchmark_rlifting$Mode == "offline" &
    benchmark_rlifting$Wavelet == "haar" &
    benchmark_rlifting$Boundary == "symmetric" &
    benchmark_rlifting$Method == "universal_semisoft",
  c("Signal", "MSE_median")
]

wt_def = benchmark_wavethresh[
  benchmark_wavethresh$Wavelet == "db1" &
    benchmark_wavethresh$Boundary == "BayesThresh_soft",
  c("Signal", "MSE_median")
]

ad_def = benchmark_adlift[
  benchmark_adlift$Wavelet == "LinearPred" &
    benchmark_adlift$Boundary == "linear_n1_int_noclo_mean",
  c("Signal", "MSE_median")
]

nlt_def = benchmark_nlt[
  benchmark_nlt$Wavelet == "LinearPred" &
    benchmark_nlt$Boundary == "linear_n1_int_noclo_mean",
  c("Signal", "MSE_median")
]

signals_ord = c("blocks", "bumps", "doppler", "heavisine")
def_wide = data.frame(
  Signal = signals_ord,
  rLifting = rl_def$MSE_median[match(signals_ord, rl_def$Signal)],
  wavethresh = wt_def$MSE_median[match(signals_ord, wt_def$Signal)],
  adlift = ad_def$MSE_median[match(signals_ord, ad_def$Signal)],
  nlt = nlt_def$MSE_median[match(signals_ord, nlt_def$Signal)]
)
knitr::kable(
  def_wide, row.names = FALSE, digits = 4,
  caption = "Regular-grid MSE at each package's default (out-of-the-box) configuration."
)
```

```{r reg-default-geom}
def_geom = data.frame(
  Package = c("rLifting", "wavethresh", "adlift", "nlt"),
  Default = round(c(
    geom_mean(def_wide$rLifting),
    geom_mean(def_wide$wavethresh),
    geom_mean(def_wide$adlift),
    geom_mean(def_wide$nlt)
  ), 4)
)
knitr::kable(
  def_geom, row.names = FALSE,
  caption = "Geometric mean MSE at default configuration (regular grid)."
)
```

At default settings, `rLifting` and `wavethresh` are essentially tied (~0.021 geomean), and `nlt` is close behind. `adlift` has the best default (0.0125) because its adaptive predictor already selects good coefficients locally — there is little the user needs to specify. On `blocks`, `rLifting`'s default (0.0116) outperforms `wavethresh`'s (0.0076 with BayesThresh; but 0.0781 with the classic `universal_soft` threshold, which shows how sensitive wavethresh is to the choice of policy). `adlift` and `nlt` with their fixed linear predictor do well on smooth signals (`doppler`, `heavisine`) but lag on `bumps`.

### Tuning cost

The gap between the default and the best achievable configuration measures how much expert knowledge each package requires:

```{r reg-tuning-cost}
best_geom = c(
  rLifting = geom_mean(reg$MSE[reg$Package == "rLifting (offline)"]),
  wavethresh = geom_mean(reg$MSE[reg$Package == "wavethresh"]),
  adlift = geom_mean(reg$MSE[reg$Package == "adlift"]),
  nlt = geom_mean(reg$MSE[reg$Package == "nlt"])
)

def_geom_v = c(
  rLifting = geom_mean(def_wide$rLifting),
  wavethresh = geom_mean(def_wide$wavethresh),
  adlift = geom_mean(def_wide$adlift),
  nlt = geom_mean(def_wide$nlt)
)

tuning_tbl = data.frame(
  Package = c("rLifting", "wavethresh", "adlift", "nlt"),
  Default = round(def_geom_v, 4),
  Best = round(best_geom,  4),
  Multiplier = round(def_geom_v / best_geom, 2)
)

knitr::kable(
  tuning_tbl, row.names = FALSE,
  caption = "Tuning cost: ratio of default to best geomean MSE. Larger = more benefit from tuning."
)
```

`adlift` needs almost no tuning (1.16×): its adaptive predictor does the heavy lifting automatically. `rLifting` is next (1.84×): its default haar + semisoft + universal is already a reasonable first guess for most signals. `wavethresh` benefits the most from tuning (2.18×): choosing the right filter family and threshold policy requires wavelet expertise. The practical implication: `rLifting` is a good package to reach for before you have signal-specific knowledge; `adlift` is the right choice when you want a self-tuning offline denoiser and speed is secondary.

## 3. Regular grid — speed

Per-sample times at each package's best configuration:

```{r reg-speed}
speed_tbl = aggregate(us ~ Package, data = reg, FUN = median)
speed_tbl$us = round(speed_tbl$us, 2)
speed_tbl = speed_tbl[order(speed_tbl$us), ]
names(speed_tbl) = c("Package", "us_per_sample_median")

knitr::kable(
  speed_tbl, row.names = FALSE,
  caption = "Per-sample time at best configuration, median across the four DJ signals."
)
```

```{r reg-speed-fig, fig.cap = "Per-sample time at each package's best configuration, log scale.", fig.height = 3.5}
ggplot(reg, aes(x = Package, y = us, fill = Package)) +
  geom_boxplot() +
  scale_y_log10() +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = expression("Per-sample time ("*mu*"s, log scale)")) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

The gap is large. `rLifting` offline is ~40× faster than `wavethresh` and ~40 000× faster than `adlift` / `nlt`. The difference is structural: `rLifting`'s offline path runs LWT + threshold + ILWT as a single C++ pass with no per-sample R allocations (`inst/notes/03-zero-allocation-engine.md`); `wavethresh` does the equivalent work in C with some R overhead per filter call; `adlift` and `nlt` implement the lifting steps in pure R closures.

Combined with §2: on the regular grid `rLifting` runs ~40× faster per sample than `wavethresh` while trailing it by ~18% at the geometric mean MSE level. On signal lengths beyond $N = 1024$ the speed gap widens linearly — the per-sample-time constant factor of the C++ engine dominates the wall time of any package once signals are long enough.

## 4. rLifting modes — regular grid

For completeness, the three rLifting modes on the same four signals:

```{r modes-mse-reg}
mode_best = function(mode_name) {
  sub = benchmark_rlifting[benchmark_rlifting$Mode == mode_name, ]
  do.call(
    rbind,
    by(
      sub, sub$Signal, 
      function(s) {
        i = which.min(s$MSE_median)
        data.frame(
          Signal = s$Signal[i], Mode = mode_name,
          MSE = s$MSE_median[i], us = s$Per_sample_us_median[i]
        )
      }
    )
  )
}

modes_reg = rbind(
  mode_best("offline"),
  mode_best("causal"),
  mode_best("stream")
)

modes_mse = reshape(
  modes_reg[, c("Signal","Mode","MSE")],
  idvar = "Signal", timevar = "Mode", direction = "wide"
)
names(modes_mse) = sub("MSE\\.", "", names(modes_mse))

modes_mse$causal_penalty = round(modes_mse$causal / modes_mse$offline, 2)
modes_mse$stream_penalty = round(modes_mse$stream / modes_mse$offline, 2)

modes_mse[, c("offline","causal","stream")] = round(
  modes_mse[, c("offline","causal","stream")],
  4
)

knitr::kable(
  modes_mse, row.names = FALSE,
  caption = "MSE per mode, regular grid. Penalty = causal/offline or stream/offline."
)
```

Sliding-window modes pay ~3–5× MSE versus offline on these signals. The penalty has two sources: the finest-level $\hat\sigma$ estimate is computed from $m_1 = W/2$ samples per window instead of $n/2$ samples globally, raising the threshold variance; and the per-window boundary zone is a larger fraction of the work than at offline scale (§6 of `vignette("04-boundary-modes")` covers the boundary side).

Per-sample times by mode:

```{r modes-speed-reg}
modes_speed = reshape(
  modes_reg[, c("Signal","Mode","us")],
  idvar = "Signal", timevar = "Mode", 
  direction = "wide"
)
names(modes_speed) = sub("us\\.", "", names(modes_speed))

modes_speed[, c("offline", "causal", "stream")] = round(
  modes_speed[, c("offline", "causal", "stream")], 2
)

knitr::kable(
  modes_speed, row.names = FALSE,
  caption = "Per-sample time per mode, regular grid (us)."
)
```

Causal is 60–130× slower per sample than offline. Causal does one full LWT + threshold + ILWT pass per sample (vs. once total for the whole signal in offline); stream pays an additional R closure invocation overhead on top of the same C++ work. Stream and causal are comparable in per-sample time; on lightweight wavelets the R closure overhead can push stream modestly above causal, while on heavier wavelets the C++ cost dominates and the two converge.

### rLifting ranking per mode

The optimal wavelet shifts between modes. The table below shows the wavelet ranking per signal at each mode (minimising over all boundaries and threshold methods):

```{r wavelet-ranking-per-mode}
wav_rank = do.call(
  rbind, lapply(
    c("offline", "causal", "stream"),
    function(mode) {
      mse_col = if (mode == "offline") "MSE_median" else "MSE_settled_median"
      sub = benchmark_rlifting[benchmark_rlifting$Mode == mode, ]
      agg = aggregate(sub[[mse_col]] ~ sub$Signal + sub$Wavelet, FUN = min)
      names(agg) = c("Signal", "Wavelet", "MSE")
      do.call(
        rbind, lapply(
          unique(agg$Signal), 
          function(s) {
            ss = agg[agg$Signal == s, ]
            ss = ss[order(ss$MSE), ]
            data.frame(
              Mode = mode,
              Signal = s,
              rank1 = ss$Wavelet[1],
              rank2 = ss$Wavelet[2],
              rank3 = ss$Wavelet[3],
              rank4 = ss$Wavelet[4]
            )
          }
        )
      )
    }
  )
)

knitr::kable(
  wav_rank, row.names = FALSE,
  caption = paste0(
    "Wavelet ranking per mode and signal (ranks 1-4 of 5 wavelets; ",
    "dd4 is consistently 5th and omitted)."
  )
)
```

Two patterns stand out. In offline mode `haar` wins only on `blocks` (a piecewise-constant signal where its single-tap predict is exact); `cdf97` or `db2` lead everywhere else. In causal and stream modes the picture inverts: `haar` dominates on `blocks`, `bumps`, and `heavisine`; `cdf97` retains its edge only on `doppler`. The reason is the sliding-window boundary zone: longer filters spend a higher fraction of the window near the edge, where boundary artefacts inflate the MSE more than the extra vanishing moments help. `dd4` is consistently last across all modes and signals — its Lagrange-4 predict step is useful on irregular grids (§7) but is not competitive on regular uniform grids where the simpler filters win.

## 5. Irregular grid — MSE

### Best achievable configuration

The question this section answers: *if your data is irregularly sampled, where does `rLifting` sit relative to the two packages built specifically for that regime?* `wavethresh` is excluded here for the reason given in §1.1 (it requires uniform sampling). On the seven irregular-grid signals, each package's best configuration:

```{r irr-mse}
irr_best = function(df, mse_col, time_col, pkg_label) {
  do.call(
    rbind, by(
      df, df$Signal, 
      function(s) {
        i = which.min(s[[mse_col]])
        data.frame(
          Signal = s$Signal[i],
          Package = pkg_label,
          MSE = s[[mse_col]][i],
          us = s[[time_col]][i] * 1e6 / s$N[i]
        )
      }
    )
  )
}

off_irr = benchmark_rlifting_irregular[
  benchmark_rlifting_irregular$Mode == "offline",
]

irr = rbind(
  irr_best(off_irr, "MSEpos_median", "Timepos_median", "rLifting (offline)"),
  irr_best(benchmark_adlift_irregular, "MSE_median", "Time_median", "adlift"),
  irr_best(benchmark_nlt_irregular, "MSE_median", "Time_median", "nlt")
)

irr_wide = reshape(
  irr[, c("Signal","Package","MSE")],
  idvar = "Signal", timevar = "Package", direction = "wide"
)
names(irr_wide) = sub("MSE\\.", "", names(irr_wide))

knitr::kable(
  irr_wide, row.names = FALSE, digits = 4,
  caption = "Irregular-grid MSE at each package's best configuration."
)
```

```{r irr-mse-fig, fig.cap = "Irregular-grid MSE at each package's best configuration, by signal.", fig.height = 4.5}
ggplot(irr, aes(x = Package, y = MSE, fill = Package)) +
  geom_col() +
  facet_wrap(~ Signal, scales = "free_y", nrow = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Median MSE") +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

`adlift` wins on all seven signals. The margin is small on six (1–30%) and large on one (`linear_phys`, where adlift is 3.2× better than rLifting). The reason for the `linear_phys` gap is the differing prediction strategy: `adlift::fwtnp` uses *adaptive* prediction, choosing predictor coefficients from local signal data; this gives near-perfect predictions on smooth signals dominated by a linear trend. `rLifting`'s cdf53 (or any of its fixed-coefficient lifting schemes) cannot match this on a signal whose first derivative is essentially constant.

`rLifting` beats `nlt` on five of seven signals; ties or loses by a few percent on the other two. Against `nlt` specifically, rLifting's lower-overhead C++ engine and SCAD shrinkage give the edge across most signal classes.

```{r irr-mse-geom}
geom_tbl_irr = aggregate(MSE ~ Package, data = irr, FUN = geom_mean)
geom_tbl_irr$MSE = round(geom_tbl_irr$MSE, 4)
geom_tbl_irr = geom_tbl_irr[order(geom_tbl_irr$MSE), ]
knitr::kable(
  geom_tbl_irr, row.names = FALSE,
  caption = "Geometric mean MSE across the seven irregular-grid signals."
)
```

By geomean, `adlift` is ~28% lower than `rLifting`, which is ~20% lower than `nlt`. Most of the adlift advantage comes from `linear_phys`; remove that signal and the geomean gap shrinks to ~10%. Within ~10% of the irregular-grid quality reference is a strong result for a package whose native design serves a wider scope (regular grids in §2, three modes in §4 and §7, sliding-window denoising at all).

### Out-of-the-box configuration

The default for each package on irregular grids: `rLifting` uses `haar` + `symmetric` boundary with physical positions `t` supplied (mandatory on irregular grids); `adlift` and `nlt` use a linear predictor with 1 neighbour, interpolating, no adaptive closure, and mean aggregation — the simplest fixed predictor available.

```{r irr-default}
rl_irr_def = benchmark_rlifting_irregular[
  benchmark_rlifting_irregular$Mode == "offline" &
    benchmark_rlifting_irregular$Wavelet == "haar" &
    benchmark_rlifting_irregular$Boundary == "symmetric" &
    benchmark_rlifting_irregular$Method == "universal_semisoft",
  c("Signal", "MSEpos_median")
]
ad_irr_def = benchmark_adlift_irregular[
  benchmark_adlift_irregular$Wavelet == "LinearPred" &
    benchmark_adlift_irregular$Boundary == "linear_n1_int_noclo_mean",
  c("Signal", "MSE_median")
]
nlt_irr_def = benchmark_nlt_irregular[
  benchmark_nlt_irregular$Wavelet == "LinearPred" &
    benchmark_nlt_irregular$Boundary == "linear_n1_int_noclo_mean",
  c("Signal", "MSE_median")
]

irr_sigs_ord = c(
  "blocks_dj_irr", "bumps_dj_irr", "doppler_dj_irr", "heavisine_dj_irr",
  "blocks_gapped", "linear_phys", "trend_events"
)

irr_def_wide = data.frame(
  Signal = irr_sigs_ord,
  rLifting = rl_irr_def$MSEpos_median[match(irr_sigs_ord, rl_irr_def$Signal)],
  adlift = ad_irr_def$MSE_median[match(irr_sigs_ord, ad_irr_def$Signal)],
  nlt = nlt_irr_def$MSE_median[match(irr_sigs_ord, nlt_irr_def$Signal)]
)

knitr::kable(
  irr_def_wide, row.names = FALSE, digits = 4,
  caption = "Irregular-grid MSE at each package's default configuration."
)
```

The default `haar` configuration splits the seven signals into two groups. On the four DJ irregular signals and `blocks_gapped` — all discontinuous or non-trending — `rLifting` is competitive and beats both alternatives on `blocks_dj_irr` (0.0103 vs adlift 0.0192) and `blocks_gapped` (0.0338 vs adlift 0.0470). On `linear_phys` and `trend_events` — signals with a strong spatial trend — the default `haar` produces catastrophically high MSE (0.45–0.50). This is expected: `haar`'s nearest-neighbour predict does not account for the local slope, so the detail coefficients are proportional to the local spacing and are not zero even for a linear signal. For trending signals, the calibrated choice is `cdf53` + `local_linear` + `t`, which recovers the physical positions and reduces detail coefficients to near-zero on linear inputs.

### Tuning cost on irregular grids

```{r irr-tuning-cost}
irr_best_geom = c(
  rLifting = geom_mean(irr$MSE[irr$Package == "rLifting (offline)"]),
  adlift = geom_mean(irr$MSE[irr$Package == "adlift"]),
  nlt = geom_mean(irr$MSE[irr$Package == "nlt"])
)

irr_def_geom = c(
  rLifting = geom_mean(irr_def_wide$rLifting),
  adlift = geom_mean(irr_def_wide$adlift),
  nlt = geom_mean(irr_def_wide$nlt)
)

irr_tuning_tbl = data.frame(
  Package = c("rLifting", "adlift", "nlt"),
  Default = round(irr_def_geom, 4),
  Best = round(irr_best_geom, 4),
  Multiplier = round(irr_def_geom / irr_best_geom, 2)
)

knitr::kable(
  irr_tuning_tbl, row.names = FALSE,
  caption = "Irregular-grid tuning cost: ratio of default to best geomean MSE."
)
```

The tuning cost for `rLifting` on irregular grids is large (driven by `linear_phys` and `trend_events`). On the five non-trending signals alone the ratio drops to ~1.7× — comparable to the regular-grid figure. The practical takeaway: if you know your signal has no strong spatial trend, `haar` + `t` is a solid starting point. If you suspect a trend, start with `cdf53` + `local_linear` + `t` instead.

## 6. Irregular grid — speed

```{r irr-speed}
irr_speed = aggregate(us ~ Package, data = irr, FUN = median)
irr_speed$us = round(irr_speed$us, 2)
irr_speed = irr_speed[order(irr_speed$us), ]
names(irr_speed) = c("Package", "us_per_sample_median")
knitr::kable(
  irr_speed, row.names = FALSE,
  caption = "Per-sample time at best configuration on irregular grids."
)
```

```{r irr-speed-fig, fig.cap = "Per-sample time on irregular grids, log scale.", fig.height = 3.5}
ggplot(irr, aes(x = Package, y = us, fill = Package)) +
  geom_boxplot() +
  scale_y_log10() +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = expression("Per-sample time ("*mu*"s, log scale)")) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

At irregular grids, `rLifting` is ~20 000× faster per sample than `adlift` and `nlt`. Both `adlift` and `nlt` perform their lifting steps in pure R with closure-heavy adaptive prediction logic; the constant-factor cost is high regardless of signal length.

This is the dimension on which the package-design choice matters most. For batch single-shot denoising of short signals where wall time is a one-second affair, the speed gap is academic. For a 100 000-sample irregular signal, or any streaming application, `rLifting` is the only package in this comparison that finishes in a useful time at all — `adlift` or `nlt` would take ~5 minutes per call where `rLifting` takes ~16 milliseconds.

The speed advantage also reframes the §5 tuning cost. For a single signal, rLifting's full offline configuration space — 360 combinations of wavelet, boundary, and threshold method — takes ~0.11 s to sweep in full. A single `adlift` or `nlt` call on the same signal takes ~3.2 s, roughly 29× longer. Running every rLifting configuration and picking the best one is still faster than a single `adlift` or `nlt` call. There is no practical trade-off between tuning and time.

## 7. rLifting modes — irregular grid

```{r modes-mse-irr}
mode_best_irr = function(mode_name) {
  sub = benchmark_rlifting_irregular[
    benchmark_rlifting_irregular$Mode == mode_name &
      !is.na(benchmark_rlifting_irregular$MSEpos_median), 
  ]
  do.call(
    rbind, by(
      sub, sub$Signal, 
      function(s) {
        i = which.min(s$MSEpos_median)
        data.frame(
          Signal = s$Signal[i], Mode = mode_name,
          MSE = s$MSEpos_median[i],
          us = s$Timepos_median[i] * 1e6 / s$N[i]
        )
      }
    )
  )
}

modes_irr = rbind(
  mode_best_irr("offline"),
  mode_best_irr("causal"),
  mode_best_irr("stream")
)

modes_mse_irr = reshape(
  modes_irr[, c("Signal","Mode","MSE")],
  idvar = "Signal", timevar = "Mode", direction = "wide"
)
names(modes_mse_irr) = sub("MSE\\.", "", names(modes_mse_irr))

modes_mse_irr$causal_penalty = round(
  modes_mse_irr$causal /modes_mse_irr$offline, 
  2
)

modes_mse_irr[, c("offline","causal","stream")] = round(
  modes_mse_irr[, c("offline","causal","stream")], 
  4
)

knitr::kable(
  modes_mse_irr, row.names = FALSE,
  caption = "MSE per mode, irregular grid. Causal/stream are essentially identical; penalty = causal/offline."
)
```

The causal-vs-offline penalty is similar in magnitude to the regular-grid case (~3–5× on most signals; ~10× on `linear_phys` where the global MAD estimate is decisive). Stream and causal share the same C++ engine and produce essentially the same MSE at any configuration; the choice between them is API ergonomics (batch over a known history vs. closure for live processing), not quality.

## 8. The speed–quality picture

A scatter of MSE against per-sample time summarises the design space:

```{r pareto, fig.cap = "Speed-quality trade-off: median MSE vs. per-sample time at each package's best configuration (regular grid, log time scale)."}
ggplot(reg, aes(x = us, y = MSE, colour = Package, shape = Package)) +
  geom_point(size = 3) +
  scale_x_log10() +
  scale_colour_brewer(palette = "Set1") +
  labs(
    x = expression("Per-sample time ("*mu*"s, log scale)"),
    y = "Median MSE"
  ) +
  facet_wrap(~ Signal, scales = "free_y") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

`rLifting` (offline) sits at the upper-left corner: lowest per-sample time, MSE within 40% of the regular-grid best on every signal (within 30% on three of four). `wavethresh` is the natural alternative when the last 10–30% of MSE matters and per-sample cost in the 1–10 μs range is acceptable. `adlift` is the choice when the absolute lowest MSE is required, in particular on irregular grids with smooth-trend structure, and the application can absorb ~20 000× more per-sample time on irregular grids (or ~40× on the regular grid). `nlt` is dominated by one of the other three on every (signal, metric) combination in this benchmark — there is no signal for which it is the right choice on these criteria.

Two capabilities that the Pareto chart cannot show because no competitor offers them at all:

- Sliding-window (causal and stream) modes on either grid. `wavethresh`, `adlift`, and `nlt` are all batch-only. The §4 and §7 numbers establish that `rLifting`'s sliding-window modes pay a 3–5× MSE penalty relative to its own offline mode. Since the offline mode already trails `adlift` and `nlt` on irregular-grid signals, causal mode sits 2–5× above the native packages' offline quality on most signals, and up to ~11–35× on smooth-trend signals (`linear_phys`) where adaptive prediction gives `adlift` a large edge. The cost of streaming is real; no competitor offers a streaming API at all. Per-sample latencies are 30–55 μs (irregular) or 4–15 μs (regular).
- Streaming-API access to the engine. `new_wavelet_stream()` returns a closure that processes a single sample per call and maintains the ring buffer between calls. This is the API surface required for live denoising of a continuous sensor stream, with no equivalent in the other three packages.

## 9. Honest limitations

The benchmark covers an opinionated slice of denoising practice. Specifically, none of the conclusions above should be extrapolated to:

- Non-Gaussian noise (Laplacian, $t$-distributed, mixture noise). The MAD-based $\hat\sigma$ in rLifting and the soft/SCAD shrinkages are robust to mild departures from Gaussianity but the relative ranking of the packages on heavy-tailed noise has not been measured here.
- Non-stationary noise across the signal. The threshold update in causal/stream modes is governed by `update_freq`, which we hold at 1; a real-world workflow may want larger values (10–50, see `vignette("03-causal-stream")` §5).
- Signals shorter or longer than $N = 1024$. Short signals (say $N < 256$) have small finest-level $d_1$ vectors and the universal threshold is poorly calibrated; long signals (say $N \geq 16384$) widen the per-sample-time gap further in favour of rLifting (since the wall time of all packages scales linearly).
- Real-world signals. The DJ classics and the seven irregular signals are deliberate canonical cases; ECG, financial tick data, geophysical traces have their own characteristic spectral content and noise structure that may shift the rankings.
- Adaptive prediction beyond what is in `adlift`. The `adlift` library uses one specific adaptive prediction strategy; richer schemes (e.g. signal-class-specific learnt predictors) may outperform every package in this comparison.

Within the slice that *is* measured, the rankings above are reliable: each cell is an average over 1 000 Monte Carlo realisations, the standard errors of the medians are small (see the `_se` columns of each dataset), and the same noise realisations were applied to each package within a row.

## 10. Decision guide

| Situation | Recommended package | Why |
|:----------|:--------------------|:----|
| Regular grid, MSE matters most, batch single-shot | `wavethresh` | lowest geomean MSE on the DJ classics; only marginally slower than rLifting offline at this scale |
| Regular grid, balanced MSE / speed | `rLifting` offline | within 10–30% of wavethresh's MSE, ~30× faster per sample |
| Regular grid, streaming or low-latency | `rLifting` causal or stream | only package in this benchmark that supports sliding-window denoising |
| Irregular grid, MSE absolutely critical (offline only) | `adlift` | wins all seven signals; up to 3× lower MSE on smooth-trend signals |
| Irregular grid, balanced MSE / speed | `rLifting` offline (with `t = t_phys`) | within ~30% of adlift's MSE on six of seven signals; 10 000× faster |
| Irregular grid, streaming or low-latency | `rLifting` causal or stream | only package in this benchmark that supports sliding-window denoising on irregular grids |
| Either grid, MSE is a soft constraint and the application is throughput-bound | `rLifting` (any mode) | the per-sample-time gap dominates everything else |

The recommendations follow the per-signal winners of §2–§7; the package-specific configuration recommendations are in `vignette("02-thresholding-and-tuning")` (regular grids), `vignette("04-boundary-modes")` (per-mode boundary effects), and `vignette("05-irregular-grids")` (irregular grids).

## 11. Where to look next

- For the underlying mechanism that lets rLifting reach these speeds, see `inst/notes/03-zero-allocation-engine.md`.
- For the empirical recommendation table specific to irregular grids, see `vignette("05-irregular-grids")` §5.
- For the cross-package raw data, see `data(benchmark_rlifting)`, `data(benchmark_wavethresh)`, `data(benchmark_adlift)`, `data(benchmark_nlt)`, `data(benchmark_rlifting_irregular)`, `data(benchmark_adlift_irregular)`, and `data(benchmark_nlt_irregular)`.

## References

Donoho, D. L., & Johnstone, I. M. (1994). Ideal spatial adaptation by wavelet shrinkage. *Biometrika*, **81**(3), 425–455.

Knight, M. I., & Nason, G. P. (2009). A 'nondecimated' lifting transform. *Statistics and Computing*, **19**(1), 1–16. [the `nlt` package]

Nason, G. P. (2008). *Wavelet Methods in Statistics with R*. Springer. [the `wavethresh` package]

Nunes, M. A., Knight, M. I., & Nason, G. P. (2006). Adaptive lifting for nonparametric regression. *Statistics and Computing*, **16**(2), 143–159. [the `adlift` package]

Sweldens, W. (1996). The lifting scheme: a custom-design construction of biorthogonal wavelets. *Applied and Computational Harmonic Analysis*, **3**(2), 186–200.
