---
title: "3. Causal decomposition: the rLifting advantage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Causal decomposition: the rLifting advantage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Most wavelet packages in R are designed for offline analysis: they assume the
entire signal is available at once. To perform causal filtering (where the
output at time $t$ depends only on data at times $0 \dots t$) with these
packages, one must implement a sliding-window loop, re-calculating the
transform at every new point.

This approach has $O(N \cdot W)$ complexity (where $W$ is the window size).
`rLifting` implements a specialized causal mode using a ring-buffer
architecture, which updates the transform state in $O(N)$ total complexity 
(where $N$ is the signal length).

As in vignette 2, the results are pre-computed by the script
`data-raw/generate_vignette_data.R`.

## 1. Speed and accuracy: optimized vs. naive

We compare `rLifting::denoise_signal_causal` against a naive sliding-window
implementation using `wavethresh`. Both methods process a HeaviSine signal
($n = 500$, $\sigma = 0.3$) using the Haar wavelet.

```{r causal_speed}
library(rLifting)

if (!requireNamespace("knitr", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Required package 'knitr' is missing. Vignette code will not run.")
} else {
  library(knitr)
}

data("benchmark_causal", package = "rLifting")

df_causal = data.frame(
  Method = c("rLifting (causal)", "Naive loop (wavethresh)"),
  Time_ms = c(benchmark_causal$rLifting_Time_Avg * 1000,
              benchmark_causal$Wavethresh_Naive_Time * 1000),
  MSE = c(benchmark_causal$rLifting_MSE,
          benchmark_causal$Wavethresh_Naive_MSE)
)

kable(df_causal,
      col.names = c("Method", "Time (ms)", "MSE"),
      digits = 4,
      caption = "Causal denoising: speed and reconstruction accuracy.")
```

`rLifting` is approximately `r round(benchmark_causal$Speedup_Factor, 0)`×
faster than the naive loop. Its MSE is also competitive, demonstrating that
the ring-buffer architecture does not sacrifice accuracy for speed.

## 2. Leakage test (look-ahead bias)

A causal filter guarantees that the output at time $t$ depends only on data
up to and including time $t$. In offline processing, the transform operates
on the entire signal at once, so a future event (such as a sudden step)
can alter the denoised output at earlier time indices through the wavelet
coefficients that span the event boundary. This is called data leakage, or
look-ahead bias.

### Methodology

To quantify leakage, we use a counterfactual test. We generate two signals
that share identical noise for $t < 128$ but differ after:

- Signal A: noise only (no event)
- Signal B: same noise, but with a step of amplitude 5 added at $t \geq 128$

We run each filter on both signals and measure the total squared difference
in their output before the step ($t < 128$):

$$\text{Leakage} = \sum_{t=1}^{127} \bigl[\hat{f}_B(t) - \hat{f}_A(t)\bigr]^2$$

A truly causal filter produces identical output for both signals in $[1, 127]$,
so its leakage is exactly zero. An offline filter may produce different output
because the step's wavelet coefficients, which have support spanning the
boundary, influence the reconstruction at earlier indices.

To make this effect visible, we use longer-support wavelets (CDF 9/7 for
`rLifting`, Daubechies-8 for `wavethresh`) with `floor(log2(N))` decomposition
levels. Longer filters have wider support, so their coefficients near the
step boundary reach further into the past.

```{r leakage_table}
data("leakage_results", package = "rLifting")

kable(leakage_results,
      col.names = c("Method", "Leakage (SSE)"),
      caption = "Counterfactual leakage: lower is better. Zero means no look-ahead bias.")
```

The causal mode of `rLifting` achieves exactly zero leakage: its output
before $t = 128$ is identical regardless of whether a step follows or not.
The filter simply has not seen the future data.

Both offline methods show non-zero leakage. The wavelet coefficients near the
step boundary ($t \approx 128$) have support that extends into the past region,
so the large discontinuity introduced by the step affects the reconstructed
values at earlier indices. This is an inherent property of offline wavelet
processing with non-trivial filter support — the longer the filter, the
further the leakage reaches.

This property is critical for financial backtesting (where look-ahead bias
leads to unrealistically optimistic results) and real-time control systems
(where the controller must not react to events that have not yet occurred).
