---
title: "2. Benchmark: speed and reconstruction accuracy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Benchmark: speed and reconstruction accuracy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)

pkgs_ok = all(
  requireNamespace("ggplot2",      quietly = TRUE),
  requireNamespace("dplyr",        quietly = TRUE),
  requireNamespace("knitr",        quietly = TRUE),
  requireNamespace("tidyr",        quietly = TRUE)
)
if (!pkgs_ok) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Required packages missing. Vignette code will not run.")
} else {
  library(ggplot2)
  library(dplyr)
  library(knitr)
  library(tidyr)
}

library(rLifting)
data("benchmark_offline", package = "rLifting")
```

This vignette benchmarks `rLifting` against `wavethresh`, `adlift`, and `nlt`
on the four classical Donoho–Johnstone (1994) test signals, evaluating both
execution speed and reconstruction accuracy (MSE). Results are pre-computed
by `data-raw/generate_vignette_data.R` and shipped as package data to comply
with CRAN time limits.

## Methodology

**Signals.** All experiments use $n = 1{,}024$ samples at noise level
$\sigma = 0.3$, following the protocol of Donoho and Johnstone (1994):

| Signal    | Character                                             |
|:----------|:------------------------------------------------------|
| Doppler   | Oscillating amplitude; challenging near $t = 0$      |
| HeaviSine | Smooth sine with two sharp discontinuities            |
| Bumps     | Narrow, irregularly spaced spikes                     |
| Blocks    | Piecewise constant with abrupt jumps                  |

**Packages and settings:**

| Package      | Paradigm            | Thresholding           | Impl.             | Sims |
|:-------------|:--------------------|:-----------------------|:------------------|-----:|
| `rLifting`   | Lifting (fixed)     | Semisoft (adaptive)    | C++ (zero-alloc)  |   50 |
| `wavethresh` | DWT                 | Universal soft         | R + C internals   |   20 |
| `adlift`     | Adaptive lifting    | EbayesThresh           | Pure R            |   20 |
| `nlt`        | Nonnested lifting   | EbayesThresh           | Pure R            |   20 |

`rLifting` results cover all five built-in wavelets (`haar`, `db2`, `cdf53`,
`cdf97`, `dd4`) and three boundary extension modes (`symmetric`,
`local_linear`, `one_sided`). The reference packages are run with their Haar
(or equivalent first-order) wavelet for a fair comparison.

---

## 1. Speed

```{r speed}
df_speed = benchmark_offline |>
  filter(!is.na(Time)) |>
  group_by(Pkg) |>
  summarise(
    Median_ms = round(median(Time) * 1e3, 3),
    Mean_ms   = round(mean(Time)   * 1e3, 3),
    .groups   = "drop"
  ) |>
  arrange(Median_ms)

kable(df_speed,
      col.names = c("Package", "Median (ms)", "Mean (ms)"),
      caption   = paste("Execution time for the full denoising pipeline",
                        "(n = 1024, all signals pooled)."))
```

`rLifting`'s zero-allocation C++ core makes it the fastest option by a wide
margin. `wavethresh` is fast but uses R with C internals. The adaptive lifting
packages (`adlift`, `nlt`) fit local polynomial predictors at every lifting
step, which is algorithmically powerful but computationally costly.

---

## 2. Reconstruction accuracy by signal

```{r mse_by_signal, fig.height = 5.5}
pkg_colors = c(
  "rLifting"   = "#009E73",
  "wavethresh" = "#0072B2",
  "adlift"     = "#E69F00",
  "nlt"        = "#CC79A7"
)

signal_labels = c(
  "doppler"   = "Doppler",
  "heavisine" = "HeaviSine",
  "bumps"     = "Bumps",
  "blocks"    = "Blocks"
)

df_mse = benchmark_offline |>
  filter(!is.na(MSE)) |>
  mutate(Signal = recode(Signal, !!!signal_labels))

ggplot(df_mse, aes(x = reorder(Pkg, MSE, median), y = MSE, fill = Pkg)) +
  geom_boxplot(outlier.size = 0.6, linewidth = 0.4) +
  facet_wrap(~Signal, scales = "free_y", nrow = 2) +
  scale_fill_manual(values = pkg_colors) +
  labs(
    title    = "Reconstruction error by signal type",
    subtitle = "n = 1024, σ = 0.3 (lower is better)",
    x = NULL, y = "MSE"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## 3. rLifting: wavelet and boundary mode

These results are unique to `rLifting`: the comparison packages do not expose
wavelet family or boundary mode choices.

```{r mse_heatmap, fig.height = 5}
wavelet_order   = c("haar", "db2", "cdf53", "cdf97", "dd4")
boundary_labels = c(
  "symmetric"    = "Symmetric",
  "local_linear" = "Local linear",
  "one_sided"    = "One-sided"
)

df_rl = benchmark_offline |>
  filter(Pkg == "rLifting", !is.na(MSE)) |>
  mutate(
    Signal   = recode(Signal, !!!signal_labels),
    Wavelet  = factor(Wavelet, levels = wavelet_order),
    Boundary = recode(Boundary, !!!boundary_labels)
  ) |>
  group_by(Signal, Wavelet, Boundary) |>
  summarise(Median_MSE = median(MSE), .groups = "drop")

ggplot(df_rl, aes(x = Wavelet, y = Boundary, fill = Median_MSE)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.4f", Median_MSE)),
            size = 2.5, colour = "white") +
  facet_wrap(~Signal, nrow = 2) +
  scale_fill_gradient(low = "#009E73", high = "#D55E00",
                      name = "Median MSE") +
  labs(
    title    = "rLifting — MSE by wavelet and boundary mode",
    subtitle = "Greener = lower error",
    x = "Wavelet", y = "Boundary mode"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r boundary_table}
df_bnd_summary = benchmark_offline |>
  filter(Pkg == "rLifting", !is.na(MSE)) |>
  mutate(Boundary = recode(Boundary, !!!boundary_labels)) |>
  group_by(Boundary) |>
  summarise(
    Median_MSE = round(median(MSE), 5),
    Best_signal = Signal[which.min(MSE)],
    .groups = "drop"
  ) |>
  arrange(Median_MSE)

kable(df_bnd_summary,
      col.names = c("Boundary mode", "Median MSE (all signals)", "Best signal"),
      caption   = "Boundary mode summary across all wavelets and signals.")
```

No boundary mode dominates in all conditions. `symmetric` provides the most
stable baseline. `local_linear` outperforms when the signal has a consistent
trend at the window edge. `one_sided` avoids inventing data beyond the edge
and performs best when no reliable local trend can be estimated.

---

## 4. Speed–accuracy trade-off

```{r tradeoff, fig.height = 4}
# For rLifting, use the best configuration per signal to avoid penalising it
# for including sub-optimal wavelet/boundary combinations.
rl_best = benchmark_offline |>
  filter(Pkg == "rLifting", !is.na(MSE)) |>
  group_by(Signal, Wavelet, Boundary) |>
  summarise(Median_MSE = median(MSE), Median_time = median(Time),
            .groups = "drop") |>
  group_by(Signal) |>
  slice_min(Median_MSE, n = 1) |>
  ungroup() |>
  summarise(Median_MSE = median(Median_MSE),
            Median_time_ms = median(Median_time) * 1e3) |>
  mutate(Pkg = "rLifting (best)")

df_trade = benchmark_offline |>
  filter(Pkg != "rLifting", !is.na(MSE), !is.na(Time)) |>
  group_by(Pkg) |>
  summarise(
    Median_time_ms = median(Time) * 1e3,
    Median_MSE     = median(MSE),
    .groups = "drop"
  ) |>
  bind_rows(rl_best)

trade_colors = c(pkg_colors, "rLifting (best)" = "#009E73")

ggplot(df_trade, aes(x = Median_time_ms, y = Median_MSE,
                     colour = Pkg, label = Pkg)) +
  geom_point(size = 3.5) +
  geom_text(nudge_y = 0.001, size = 3.2) +
  scale_x_log10() +
  scale_colour_manual(values = trade_colors) +
  labs(
    title    = "Speed vs accuracy trade-off (all signals, log scale)",
    subtitle = "Lower-left is better. rLifting shown at best config per signal.",
    x = "Median execution time in ms (log scale)",
    y = "Median MSE"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

`adlift` achieves the best median MSE by adapting the predictor at every
lifting step across the entire signal; the cost is roughly 40 000× slower
execution than `rLifting`. `nlt` is comparable in speed to `adlift` but
slightly less accurate. `rLifting`'s best configuration (CDF 9/7,
`symmetric`) matches or surpasses `adlift` on smooth signals (Doppler) while
running at sub-millisecond speed — a practically relevant trade-off when
real-time or large-scale processing is required.
